Performance-portable Parallel Programming in Computer Science Education with Kokkos

### 2020-01-24 src/2.GenericProgramming.tex
first three paragraph edits

A programming model exposes constrained semantics to the programmer which the programmer uses to express intent. Absent an embedded programming model, the semantics of C-like languages can lead to sequential thinking and execution. Parallel programming models expose parallel semantics to enable developers to create performance portable code. This is commonly achieved by adding semantic value to the base programming language, often referred to as programming with \emph{pragma} annotations. Other methods expose such semantics through library calls or new languages. All three approaches are conceptually suitable to represent programming paradigms such as tasking, parallel patterns and others. Some representatives are the OpenMP programming model~\cite{CITEOPENMP}, threading libraries, or languages like Erlang in which parallelism is a first-class citizen. In general, it is up to the parallel programming model to expose language constructs or library interfaces (APIs) such that they combine convenience to the programmer with semantics sufficient to map the program execution to the parallel computer hardware efficiently and correctly. Questions of which semantic information is needed, what paradigm to present and by what syntactical means span an exciting design space worth exploring.

@@

\begin{figure}[h]
\begin{Verbatim}[frame=leftline]
Constrained Semantics: memory allocations, memory movement, parallel kernels
Parallel Paradigms: patterns, tasks
Expression: language 
Concern: prescriptive, descriptive
\end{Verbatim}
\caption{Semantic capture - a set of constrained semantics, parallel paradigms, expression and concern define key properties a parallel programming model.}
\label{figSemCapture}
\end{figure}

Constrained semantics provided by the programmer to the parallel programming model can be grouped into information to express intent and properties. They address the question of \emph{what}, \emph{where} and \emph{how}. In the context of parallel programming, these correspond to defining which code portion to parallelize, where to run and access and how to run that parallel code. Synchronization primitives may be considered as part of the what and information on the execution properties such as data placement or memory access type as the how. Semantics are expressed following a paradigm and a programming language.

The parallel programming paradigm is an abstract ~\emph{representation} with the purpose of facilitating the programmer's understanding of programming rules and program behavior. Which programming paradigm to chose depends on several considerations. Parallel-patterns allow to express concurrency for commonly occurring programming patterns such as loop constructs. Tasking is a paradigm that support the expression of concurrent loops as well as irregular algorithms. Distributed and correctness-oriented programming models may implement actor-based programming. In this programming model, each unit of execution represent an actor who communicates over predefined communication interfaces. This paradigm eliminates accesses to shared state and align well with message passing programming (MPI). The execution model and memory model of a programming model defines behavior. That is, it defines the relationship between abstract concepts and program behavior on the given architecture.

>>>>

A programming model exposes constrained semantics that the programmer uses to express intent. Absent a programming model providing additional constraints, the semantics of C-like languages can lead to sequential thinking and execution. Parallel programming models expose parallel semantics to enable developers to create performance-portable code. This is commonly achieved by adding annotations to the base programming language, often achieved using \emph{pragma} clauses or similar language extensions. Other methods expose such semantics through library calls or new languages. All three approaches are conceptually suitable to represent programming paradigms such as tasking, data parallelism, and other patterns. Some representatives are the OpenMP programming model~\cite{CITEOPENMP}, threading libraries, or languages like Erlang in which parallelism is a first-class citizen. In general, it is up to the parallel programming model implementer to expose language constructs or library interfaces (APIs) that combine convenience to the programmer with sufficiently constrained semantics to map the program execution to the parallel computer hardware efficiently and correctly. Questions of which semantic information is needed, what paradigm to present, and by what syntactic means to do so span an exciting design space that is well worth exploring.

@@

\begin{figure}[h]
\begin{Verbatim}[frame=leftline]
Constrained Semantics: memory allocations, memory movement, parallel kernels
Parallel Paradigms: patterns, tasks
Expression: language 
Control Paradigm: prescriptive, descriptive
\end{Verbatim}
\caption{Semantic capture: a set of constrained semantics, parallel paradigms, expression and concern define key properties a parallel programming model.}
\label{figSemCapture}
\end{figure}

\emph{Constrained semantics} provided by the programmer to the parallel programming model can be grouped into information to express intent or correctness and properties that control critical performance aspects. They address the question of \emph{what}, \emph{where} and \emph{how}. In the context of parallel programming, these correspond to defining which code portion to parallelize, where to run the code, where to access the data, and how to run that parallel code. Synchronization primitives may be considered part of the ``what,'' and information on the execution properties such as data placement or memory access type may be considered as part of the ``how.'' Application logic is expressed following a \emph{paradigm} implemented in a \emph{programming language}.

A parallel programming paradigm is an abstract representation with the purpose of facilitating the programmer's understanding of programming rules and program behavior. Which programming paradigm to chose depends on several considerations. Parallel patterns allow the programmer to permit concurrent execution for commonly occurring programming patterns such as loop constructs. Tasking is a paradigm that supports the expression of interdependent work items needed for irregular algorithms. Distributed and correctness-oriented programming models may implement actor-based programming, where each unit of execution represents an actor who communicates over predefined communication channels. This paradigm eliminates accesses to shared state and aligns well with message passing programming (MPI). The execution model and memory model aspects of a programming model define its behavior. That is, it defines the relationship between abstract concepts and program execution on the given architecture.
